Formal Modelling of a Microcontroller Instruction Set in B

### 2009-07-01 papers/SBMF2009/source/paper.tex
Final submetido

The B method~\cite{Abrial} supports the construction of safety systems
model by verification of proofs that guarantees its correctness. So,
an initial abstract model of the system requirements is defined and
then it is refined until the implementation model. Development
environments based on the B method also include source code generators
for several programming languages, but the result of this translation
cannot be compared by formal means. \cite{Dantas_SBMF08} presented
recently an approach to extend the scope of the B method up to the
assembly level language language. One key component of this approach
is to build, within the framework of the B method, formal modelos of
the instruction set of such assembly languages.

@@

\section{B Method}
\label{sec:B_method}

The B method for software development~\cite{Abrial} is based on the B
Abstract Machine Notation (AMN) and the use of formally proved
refinements up to a specification sufficiently concrete that
programming code can be automatically generated from it. Its
mathematical basis consists of first order logic, integer arithmetic
and set theory, and its corresponding constructs are very similar to
those of the Z notation

A B specification is structured in modules. A module defines a set of
valid states, including a set of initial states, and operations that
may provoke a transition between states. The design process starts
with a module with a so-called functional model of the system under
development. In this initial modelling stage, the B method requires
that the user proves that, in a machine, all the initial states are
valid, and that operations do not define transitions from valid states
to invalid states.

Essentially, a B module contains two main parts: a header and the
available operations. Figure~\ref{fig:maqB} has a very basic
example. The clause $\mathit{MACHINE}$ has the name of module.  The
following two clauses respectively reference external modules and
create an instance of an external module. The $\mathit{VARIABLES}$
declares the name of the variables that compose the state of the
machine. Next, the $\mathit{INVARIANT}$ clause defines the type and
other restrictions on the variables. The $\mathit{INITIALIZATION} $
sepecifies the initial states. Finally, operations correspond to
the transitions between states of the machine.

@@

We have been developed a reusable set of basic definitions to model
hardware concepts and data types concepts. These definitions are
grouped into two separate development projects and are available as
libraries. A third project is devoted to the higher-level aspects of
the Z80 model. Thus, the workspace is composed of: a hardware library,
a types library and a project for the specific platform, in this case
the Z80. The corresponding dependency diagram is depicted in
Figure~\ref{fig:hardware-definition-graph}; information specific to
each project is presented in the following.

\begin{figure}[h] \centering
\includegraphics[width=.92\textwidth]{diagramaEstrutural_vertical.jpg}
 \caption{Dependency diagram of the Z80 model.}
\label{fig:hardware-definition-graph}
\end{figure}

\subsection{Definitions to Represent and Manipulate Bits}

@@

The $\mathit{BYTE\_INDEX}$ is domain of byte modelled. It starts at 1 to obey a definition of sequences from B
method. However, the functions from byte encapsulate the access and these functions use the
$\mathit{PHYS\_BYTE\_INDEX}$. The $\mathit{BYTE}$ type is a specialized type from $\mathit{BIT\_VECTOR}$, but it has
a size limit. Some others more specific definitions are created to help the developer. The type $\mathit{BV16}$ is
created for bit vector of length 16 in a similar way.

@@

The \textit{Z80} is a CISC microcontroller developed by
\textit{Zilog}~\cite{Z80_manual}. It supports 158 different
instructions. These instructions are classified into these categories:
load and exchange; block transfer and search; arithmetic and logical;
rotate and shift; bit manipulation (set, reset, test); jump, call and
return; input/output; and basic cpu control.

The main module includes an instance of memory module and accesses the definitions from: basic data types modules
and \textit{ALU}.

@@

% Chacteristics of microcontroller
Each instruction is represented by B operations in the module Z80. By default, all parameters from
operations are or predefined elements in the model or integers values in the decimal representation.
The internal registers contain 208 bits of reading/writing memory. It includes two sets of six general purpose
registers which may be used individually as 8-bits registers or as 16-bits register pairs.  The work registers are
represented by variable $\mathit{rgs8}$. The domain of $\mathit{rgs8}$ ($\mathit{id\_regs8}$) is a set formed by
identifiers of registers of 8 bits. These registers can be accessed in pairs, forming 16-bits, resulting in other
set of identifiers of 16-bits registers, named $\mathit{id\_reg16}$. The main work register of Z80 is the
accumulator ($\mathit{rgs8(a0)}$) used for arithmetic, logic, input/output and loading/storing operations.
\subsection{Modelling registers, input and output ports and instructions}

The Z80 has many register of differents types and many instructions. It includes alternative set of accumulator,
flag and general registers. The CPU contains a stack pointer ($\mathit{sp}$), program counter ($\mathit{pc}$), two
index registers ($\mathit{ix}$ and $\mathit{iy}$), an interrupt register ($\mathit{i\_}$), a refresh register
($\mathit{r\_}$), two bits ($\mathit{iff1}$, $\mathit{iff2}$) used to control the interruptions, a pair of bits to
define the interruption mode ($\mathit{im}$) and the input and output ports ($\mathit{i\_o\_ports}$). Below, its
definitions are specified in the \textit{INVARIANT}.

\begin{sloppypar}
\bf INVARIANT

\hspace*{0.10in}\it rgs8  $\in$  \it id\_reg\_8  $\fun$  \it BYTE  $\land$ pc  $\in$  \it INSTRUCTION  $\land$

@@

 The proof obligations allow to verify the data types, important system properties and if the expressions are
 well-defined (WD)\footnote{An expression is called ``well-defined'' (or unambiguous) if its definition assigns it a
 unique interpretation or value.}. The properties provide additional guarantees, because it can set many safety
 rules. However, the model can be very difficult to prove.

% [Conclusao] This project has many proof obligations and
The building of model was a hard and difficult task, because we needed to build, prove, fix and re-prove
several predicates of the libraries.

However, few proof commands\footnote{The proof commands are step that just indicate a walk to prover make the proof,
thus theses commands cannot introduce false hypotheses.} can be useful to prove the majority of the proof
obligations. Because, there are many similar assembler instructions. A good example is a set of 17 proof commands
that quickly aided the verification of 99\% (2295) of proofs WD. Finally, the techniques mentioned  help
the verification of all the 2926 proof obligations.

@@

There are in the literature of computer science some
approaches~\cite{BHDL_2003,GEMPLUS_99} to model hardware and the
virtual machines using the B method. Then, the B method has been used
successfully to model the operational semantic. However the cost of
modelling was still expensive and this paper quoted some techniques to
lower the cost of modelling. Although, the modelling of the Z80 do not
have modelled some small aspects.
% The objective of researchers

In general, the researchers~\cite{GEMPLUS_99} are concentrated at
model the Java Virtual Machines (JVM), but it is not easy to specify,
because it is very big and complex. Then, this work specify a model
more simple to get better results, at the same time, the model stay
near to model of actual microcontrollers. The main motivation of our
research is the verification of assembler level and consequently the
verification of hardware model. Thus, the some aspects are not
modelled. For example, the time of execution of instruction, the
pipeline and others questions not directly related with the semantic
software. However, there are many other specialized techniques to
verify these questions.

@@

This work has shown an approach to the formal modelling of the
instruction set of microcontrollers using the B method. During the
construction of this model, several ambiguities and errors were
encountered in the during the official reference~\cite{Z80_manual}. As
the B notation has a syntax that is not too distant from that of
imperative programming languages, such model could be used to improve
the documentation used by assembler programmers. Besides, the formal
notation used is analyzed by software that guarantees the
correctness of typing, the well-definedness of expressions, in addition
to safety properties of the micro-controller state.

@@

\paragraph{Acknowledges:}
This work received support from ANP (\textit{Agência Nacional do
  Petróleo, Gás Natural e Biocombustíveis}) and CNPq.

>>>>

The B method~\cite{Abrial} supports the construction of safety systems
models by verification of proofs that guarantees its correctness. So,
an initial abstract model of the system requirements is defined and
then it is refined until the implementation model. Development
environments based on the B method also include source code generators
for programming languages, but the result of this translation
cannot be compared by formal means. The paper \cite{Dantas_SBMF08} presented
recently an approach to extend the scope of the B method up to the
assembly level language. One key component of this approach
is to build, within the framework of the B method, formal models of
the instruction set of such assembly languages.

@@

\section{Introduction to the B Method}
\label{sec:B_method}

The B method for software development~\cite{Abrial} is based on the B Abstract
Machine Notation (AMN) and the use of formally proved refinements up to a
specification sufficiently concrete that programming code can be automatically
generated from it. Its mathematical basis consists of first order logic, integer
arithmetic and set theory, and its corresponding constructs are similar to those
of the Z notation.

A B specification is structured in modules. A module defines a set of valid
states, including a set of initial states, and operations that may provoke a
transition between states. The design process starts with a module with a
so-called functional model of the system under development. In this initial
modelling stage, the B method requires that the user proves that, in a machine,
all the its initial states are valid, and that operations do not define
transitions from valid states to invalid states.

Essentially, a B module contains two main parts: a header and the available
operations. Figure~\ref{fig:maqB} has a very basic example. The clause
$\mathit{MACHINE}$ has the name of module.  The next two clauses respectively
reference external modules and create an instance of an external module. The
$\mathit{VARIABLES}$ clauses declares the name of the variables that compose the
state of the machine. Next, the $\mathit{INVARIANT}$ clause defines the type and
other restrictions on the variables. The $\mathit{INITIALIZATION} $ specifies
the initial states. Finally, operations correspond to the transitions between
states of the machine.

@@

We have been developed a reusable set of basic definitions to model
hardware concepts and data types concepts. These definitions are
grouped into two separate development projects and are available as
libraries. A third project is devoted to the higher-level aspects of
the platform. Thus, the workspace is composed of: a hardware library,
a types library and a project for the specific platform, in this case
the Z80. The corresponding dependency diagram is depicted in
Figure~\ref{fig:hardware-definition-graph}; information specific to
each project is presented in the following.

\begin{figure}[h] \centering
\includegraphics[width=.92\textwidth]{diagramaEstrutural.png}
 \caption{Dependency diagram of the Z80 model.}
\label{fig:hardware-definition-graph}
\end{figure}

\subsection{Bit Representation and Manipulation}

@@

The $\mathit{BYTE\_INDEX}$ is the domain of the functions modelling bytes. It
starts at 1 to obey a definition of sequences from B. However, it is common in
hardware architectures to start indexing from zero. The definition
$\mathit{PHYS\_BYTE\_INDEX}$ is used to provide functionalities obeying this
convention. The $\mathit{BYTE}$ type is a specialized type from
$\mathit{BIT\_VECTOR}$, but it has a size limit. Other specific definitions are
provided to facilitate further modelling: the type $\mathit{BV16}$ is created for
bit vector of length 16 in a similar way.

@@

The \textit{Z80} is a CISC microcontroller developed by
\textit{Zilog}~\cite{Z80_manual}. It supports 158 different instructions and all
of them were specified. These instructions are classified into these categories:
load and exchange; block transfer and search; arithmetic and logical; rotate and
shift; bit manipulation; jump, call and return; input/output; and basic cpu
control.

The main module includes an instance of the memory module and accesses the
definitions from basic data types modules and the \textit{ALU} module.

@@

% Chacteristics of microcontroller
Each instruction is represented by a B operation in the module Z80. By
default, all parameters from operations are either predefined elements
in the model or integers values in the decimal representation.  The
internal registers contain 208 bits of reading/writing memory. It
includes two sets of six general purpose registers which may be used
individually as 8-bits registers or as 16-bits register pairs.  The
working registers are represented by variable $\mathit{rgs8}$. The
domain of $\mathit{rgs8}$ ($\mathit{id\_regs8}$) is a set formed by
identifiers of registers of 8 bits. These registers can be accessed in
pairs, forming 16-bits, resulting in another set of identifiers of
16-bits registers, named $\mathit{id\_reg16}$. The main working
register of Z80 is the accumulator ($\mathit{rgs8(a0)}$) used for
arithmetic, logic, input/output and loading/storing operations.

\subsection{Modelling Registers, Input and Output Ports and Instructions}

The Z80 has different types of registers and instructions. The CPU
contains general-purpose registers ($\mathit{id\_reg\_8}$), a stack
pointer ($\mathit{sp}$), program counter ($\mathit{pc}$), two index
registers ($\mathit{ix}$ and $\mathit{iy}$), an interrupt register
($\mathit{i\_}$), a refresh register ($\mathit{r\_}$), two bits
($\mathit{iff1}$, $\mathit{iff2}$) used to control the interruptions,
a pair of bits to define the interruption mode ($\mathit{im}$) and the
input and output ports ($\mathit{i\_o\_ports}$). Below, part of the
corresponding definitions are replicated from the $\INVARIANT$:

\begin{sloppypar}
%\bf INVARIANT
\hspace*{0.10in}\it rgs8  $\in$  \it id\_reg\_8  $\fun$  \it BYTE  $\land$ pc  $\in$  \it INSTRUCTION  $\land$

@@

The proof obligations allow to verify the data types, important system
properties and if the expressions are well-defined (WD)\footnote{An
  expression is called ``well-defined'' (or unambiguous) if its
  definition assigns it a unique interpretation or value.}. The
properties provide additional guarantees, because they can set many
safety rules. However, the model can be very difficult to prove.

% [Conclusao] This project has many proof obligations and
% The construction of then model was a difficult task, as several
% iterations were needed to provide the good library definitions 
% as well as to optimize the definition of the functionality of
% the microprocessor instructions by grouping common data
% manipulation into auxiliary functions.

Several iterations were needed to provide the good library definitions as well as
to fine-tune the model of the microcontroller instructions by factoring common
functionalities into auxiliary definitions.

However, few proof commands\footnote{The proof commands are steps that direct the
prover to find the proof, and cannot introduce false hypothesis.} need to be used
to prove most proof obligations. As there are many similar assembly instructions,
some human-directed proofs, when replayed, could discharge other proof
obligations. A good example is a set of 17 proof commands that quickly aided the
verification of 99\% (2295) of WD proofs. We also set up a proving environment
consisting of networked computers to take advantage of the distribution
facilities now provided in the B development environment. Finally, all of the
2926 proof obligations were proved using the tool support of the development
environment.

@@

There are in the literature of computer science some
approaches~\cite{BHDL_2003,GEMPLUS_99} to model hardware and the virtual machines
using the B method. Then, in both works the B method has been used successfully
to model the operational semantic. However the cost of modelling was still expensive and this
paper quoted some techniques to lower the cost of modelling.
% Although, the modelling of the Z80 do not have modelled some small aspects.

% The objective of researchers

In general, the researchers employing the B method have focused on
more abstract level of description of software.  Considering low-level
aspect, there has been previous work on modelling the Java Virtual
Machine~\cite{GEMPLUS_99}.

The main motivation of our research is the development of verified
software up to the assembly level, which requires specifying the
semantics of the underlying hardware. Thus, some aspects were not
modelled in our work such as the execution time of the instructions.
Also we did not consider the microarchitecture of the hardware as the scope of
our work does not include hardware verification. However, there are
many other specialized techniques to verify these questions.

@@

This work has shown an approach to the formal modelling of the instruction set of
microcontrollers using the B method. During the construction of this model, some
ambiguities and errors were encountered in the official reference for Z80
microcontroller~\cite{Z80_manual}. As the B notation has a syntax that is not too
distant from that of imperative programming languages, such model could be used
to improve the documentation used by assembler programmers. Besides, the formal
notation used is analyzed by software that guarantees the correctness of typing,
the well-definedness of expressions, in addition to safety properties of the
microcontroller state.

@@

\paragraph{Acknowledges:}
This work received support from ANP (\textit{Ag\^{e}ncia Nacional do
  Petr\'{o}leo, G\'{a}s Natural e Biocombust\'{\i}veis}) and CNPq
(\textit{Conselho Nacional de Desenvolvimento Cient\'{\i}fico e
  Tecnol\'{o}gico}).
